# chapter1: Descriptive statistics

In this chapter we will learn descriptive statistics.

## Measures of central tendency

Measures of Central Tendency are statistical values that represent the *center* or *typical value* of a data set. They help us understand the overall trend or “average” behavior of the data.

### Mode in R

The mode is the value(s) that appear most frequently in the dataset.

In R, the `lsr` package provides a simple and clean way to calculate the mode.

Step 1: Install the `lsr` Package

```{r}
install.packages("lsr")
```

Step 2: Load the Package

```{r}
library(lsr)
```

Example

```{r}
data <- c(1, 2, 2, 3, 3, 3, 4, 5)
modeOf(data)
```

⚠️ Important Note

There is a base R function also named `mode()` — but it returns the data type (like `"numeric"`, `"character"`) instead of calculating the statistical mode!

Mode in Data Frame

```{r}
df <- data.frame(
  Score = c(1, 2, 2, 3, 4, 3, NA),
  Group = c("A", "B", "B", "B", "A", "A", NA),
  Age   = c(20, 22, 21, 22, 21, 22, 21)
)

sapply(df, modeOf)
```

The `sapply()` function is executed on each column of the dataframe.

### Mode in Python

Calculating the Mode in Python Using `pandas`

Step 1: Import `pandas`

```{python}
import pandas as pd
```

**Step 2:** Create a Sample DataFrame

```{python}
data = {
    "Name": ["Alice", "Bob", "Charlie", "Diana", "Eli", "Fiona"],
    "Age": [24, 30, 22, 24, 30, 24],
    "Gender": ["Female", "Male", "Male", "Female", "Male", "Female"]
}

df = pd.DataFrame(data)
df
```
