# chapter2: Measures of variability

**Variability is most commonly measured with the following descriptive statistics:**

-   Range: the difference between the highest and lowest values.

-   Interquartile range: the range of the middle half of a distribution.

-   Standard deviation: average distance from the mean.

-   Variance: average of squared distances from the mean.

## Range in R

To calculate the range, simply subtract the largest data value from the smallest value.

```{r}
x <- c(12, 15, 18, 25, 30, 35)

range_value <- max(x) - min(x)
range_value
```

**Using range() function**

```{r}
x <- c(12, 15, 18, 25, 30, 35)
range(x)
```

**Using range() function with `NA`**

```{r}
x <- c(10, 15, NA, 25, 30, NA, 40)

range(x, na.rm = TRUE)
```

**Range in DataFrame**

```{r}
df <- data.frame(
  A = c(10, 15, 20, 25, 30),
  B = c(5, 7, 9, 12, 15),
  C = c(100, 120, 110, 130, 125),
  Gender = c("m", "f", "f", "m", "f")
)

df
```

**Using: `sapply` Numeric column**

```{r}
range_df <- sapply(df[, sapply(df, is.numeric)], function(x) max(x) - min(x))
range_df
```

**`aggregate()`** Categorical variables

```{r}
aggregate(cbind(A, B, C) ~ Gender, data = df,
          FUN = function(x) max(x) - min(x))
```

**Range in DataFrame with `NA`**

```{r}
df <- data.frame(
  A = c(10, 15, 20, 25, 30),
  B = c(5, NA, 9, 12, 15),
  C = c(100, 120, 110, 130, NA),
  Gender = c("m", "f", "f", "m", "f")
)

df
```

In R, to handle missing data we need to put `na.rm = TRUE` inside the function:

```{r}
aggregate(cbind(A, B, C) ~ Gender, data = df,
          FUN = function(x) max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
```

## **Deciles, Percentiles, and Quartiles in R**

### **1.Deciles in R**

Deciles divide the data into 10 equal groups (each 10%).

```{r}
data <- c(12, 5, 7, 9, 15, 18, 22, 25, 30, 35, 40, 50)
quantile(data, probs = seq(0, 1,by = 0.1))
```

### 2. **Percentiles in R**

To calculate any percentile, use the same function with a decimal value.

**For exampl**e, the 25th percentile and the 90th percentile:

```{r}
data <- c(12, 5, 7, 9, 15, 18, 22, 25, 30, 35, 40, 50)
quantile(data, probs = c(0.25, 0.9))

```

Or all percentiles (as a complete vector):

```{r}
data <- c(12, 5, 7, 9, 15, 18, 22, 25, 30, 35, 40, 50)
quantile(data, probs = seq(0, 1, 0.01))
```

### 3. **Quartiles in R**

You can use the `quantile()` function to compute quartiles.

```{r}
data <- c(12, 5, 7, 9, 15, 18, 22, 25, 30, 35, 40, 50)
quantile(data)

quantile(data, probs = c(0.25, 0.5, 0.75))
```

**Quartiles with `NA`**

`na.rm = TRUE`

```{r}
data <- c(12, 5, 7, NA, 15, 18, 22, NA, 30, 35, 40, 50)
quantile(data, probs = c(0.25, 0.5, 0.75), na.rm = TRUE)
```

If you want, without removing `NA` but replacing it with the `mean`.

```{r}
data <- c(12, 5, 7, NA, 15, 18, 22, NA, 30, 35, 40, 50)
data_filled <- ifelse(is.na(data), mean(data, na.rm = TRUE), data)
quantile(data_filled)
```

**DataFrame with `NA`**

Calculating quartiles for each column

```{r}
df <- data.frame(
  height = c(160, 165, 170, NA, 180, 190, NA),
  weight = c(50, 55, NA, 70, 80, 90, 95),
  age    = c(20, 25, 30, 35, NA, 45, 50)
)

sapply(df, function(x) quantile(x, probs = c(0.25, 0.5, 0.75), na.rm = TRUE))
```

**for a column**

```{r}
quantile(df$weight, probs = c(0.25, 0.5, 0.75), na.rm = TRUE)
```

**DataFrame with Categorical variable**

```{r}
df <- data.frame(
  height = c(160, 165, 170, NA, 180, 190, NA),
  weight = c(50, 55, NA, 70, 80, 90, 95),
  age    = c(20, 25, 30, 35, NA, 45, 50),
  Gender = c('Male', 'Female', 'Male', 'Female', 'Male', 'Female', 'Male')
)

quantile(df$height, probs = c(0.25, 0.5, 0.75), na.rm = TRUE)
quantile(df$weight, probs = c(0.25, 0.5, 0.75), na.rm = TRUE)
quantile(df$age,    probs = c(0.25, 0.5, 0.75), na.rm = TRUE)

```

**with `dplyr`**

```{r}
library(dplyr)

df %>%
  group_by(Gender) %>%
  summarise(
    height_Q1 = quantile(height, 0.25, na.rm = TRUE),
    height_Median = quantile(height, 0.5, na.rm = TRUE),
    height_Q3 = quantile(height, 0.75, na.rm = TRUE),
    
    weight_Q1 = quantile(weight, 0.25, na.rm = TRUE),
    weight_Median = quantile(weight, 0.5, na.rm = TRUE),
    weight_Q3 = quantile(weight, 0.75, na.rm = TRUE),
    
    age_Q1 = quantile(age, 0.25, na.rm = TRUE),
    age_Median = quantile(age, 0.5, na.rm = TRUE),
    age_Q3 = quantile(age, 0.75, na.rm = TRUE)
  )

```

**with `tapply`**

**By Gender variable**

```{r}
df <- data.frame(
  height = c(160, 165, 170, NA, 180, 190, NA),
  weight = c(50, 55, NA, 70, 80, 90, 95),
  age    = c(20, 25, 30, 35, NA, 45, 50),
  Gender = c('Male', 'Female', 'Male', 'Female', 'Male', 'Female', 'Male')
)

tapply(df$height, df$Gender, function(x) quantile(x, probs = c(0.25,0.5,0.75), na.rm = TRUE))

```

## Interquartile Range in R - IQR

the interquartile range is the difference between the upper quartile and the lower quartile.

```         
IQR=Q3−Q1
```

**`IQR()`** **Function**

```{r}
data <- c(12, 5, 7, 9, 15, 18, 22, 25, 30, 35, 40, 50)
IQR(data)
```

**Function `quantile()`**

```{r}
Q1 <- quantile(data, 0.25)
Q3 <- quantile(data, 0.75)
IQR <- Q3 - Q1
IQR
```

**DataFrame (Numeric)**

**`sapply`**

```{r}
df <- data.frame(
  height = c(160, 165, 170, 172, 180, 190),
  weight = c(50, 55, 60, 70, 80, 90)
)

sapply(df, IQR)
```

**DataFrame with `NA`**

```{r}
df <- data.frame(
  height = c(160, 165, 170, NA, 180, 190, NA),
  weight = c(50, 55, NA, 70, 80, 90, 95),
  age    = c(20, 25, 30, 35, NA, 45, 50)
)

sapply(df, IQR, na.rm = TRUE)
```

Just for weight

```{r}
IQR(df$weight, na.rm = TRUE)
```

**DataFrame (Numerical + Categorical) variable**

```{r}
df <- data.frame(
  gender = c("Male", "Female", "Male", "Female", "Male", "Female", "Male"),
  height = c(160, 165, 170, NA, 180, 190, NA),
  weight = c(50, 55, NA, 70, 80, 90, 95),
  age    = c(20, 25, 30, 35, NA, 45, 50)
)
df

by(df$weight, df$gender, function(x) quantile(x, probs = c(0.25, 0.5, 0.75), na.rm = TRUE))

```

```{r}
tapply(df$weight, df$gender, IQR, na.rm = TRUE)
```

If you want to calculate height, weight, and age all at once:

```{r}
library(dplyr)

df %>%
  group_by(gender) %>%
  summarise(
    IQR_height = IQR(height, na.rm = TRUE),
    IQR_weight = IQR(weight, na.rm = TRUE),
    IQR_age    = IQR(age, na.rm = TRUE)
  )
```

## Mean absolute deviation in R

**What is Mean Absolute Deviation?**

It is the **average of the absolute differences** between each value and the **mean** of the data.

So the steps are:

1.  Compute the mean of the data.

2.  Subtract the mean from each value (deviation).

3.  Take the absolute value of each deviation.

4.  Take the mean of those absolute values.

5.  The `lsr` package does all of this for you with one simple function: **`aad()`**.

```{r}
install.packages("lsr")
library(lsr)

scores <- c(10, 12, 13, 15, 18)
aad(scores)
```

This will:

-   compute the mean of `scores`,

-   compute the absolute deviations from that mean,

-   return the **average** of those absolute deviations.

-   So `aad(scores)` is equivalent to:

```{r}
mean(abs(scores - mean(scores)))
```

**Comparing with the base R `mad()` function**

Be careful: base R already has a function named `mad()`:

But:

-   `mad()` = **Median Absolute Deviation** (based on the **median** and usually scaled)

-   `aad()` (from `lsr`) = **Average Absolute Deviation** (typically from the **mean**, unless you specify a different center).

**Working with data with missing values:**

```{r}
library(lsr)

data <- c(12, 15, NA, 22, 25, 28, NA, 35, 40, 45)
aad(data, na.rm = TRUE)
```

**DataFrame**

```{r}
df <- data.frame(
  age = c(25, 30, 35, 40, 45, 28, 32, 38, 42, 50,
          22, 27, 33, 36, 44, 29, 31, 39, 41, 47),
  salary = c(30000, 35000, 42000, 50000, 58000, 
             32000, 38000, 45000, 52000, 60000,
             28000, 33000, 40000, 47000, 55000,
             31000, 37000, 44000, 51000, 59000),
  Gender = c("Male", "Female", "Male", "Female", "Male",
             "Female", "Male", "Female", "Male", "Female",
             "Male", "Female", "Male", "Female", "Male",
             "Female", "Male", "Female", "Male", "Female")
)
```

**Numerical column**

```{r}
aad(df$age)

aad(df$salary)
```

**MAD by Gender**

```{r}
tapply(df$age, df$Gender, aad)
tapply(df$salary, df$Gender, aad)
```
